Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-06-18 manuscript/manuscript.Rmd
add section 2.4

@@

(ref:descr-fig1-caption) Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B). Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (< 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (>500 ms) are controlled by the target stimulus.

@@

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus whose presence (s)he has to detect by pressing a button within a certain time window (e.g., the first second after target onset). In EHA parlance, the single event of interest is a button press response, *time zero* is defined as target display onset, the  *observation period* is 1 second long in each trial or repeated measurement, in each trial the participant is *at risk* for response occurrence as long as the response has not occurred yet, and the individual starts in an "idle" state in each trial and *transitions* to a "detected" state when a response occurs.

@@

After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins; Figure 1B), let RT be a discrete random variable denoting the *rank* of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the detection response in trial 1 might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).

Interval-censored... +

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence

and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in section A of the Supplemental Material.

def hazard

def survival

def cumulative distribution

def prob mass function

def conditional accuracz function

The discrete-time hazard function for individual i in trial j of condition k gives you, for each time bin, the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins.

In other words, it reflects the instantaneous risk that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the survival probability, the probability that the event occurs after bin t. In other words, the survivor function gives you for each time bin the likelihood that the event occurs in the future, i.e., in one of the subsequent time bins.

 life table...

## 2.5 Number of samples, repeated measures, time bins

In a typical RT data set from a within-subject design, there are N individuals and M repeated measures or trials per experimental condition.

Power

@@

bin width

## 2.6 Bayesian vs. frequentist approach

To study how the risk of a response, and the accuracy of an emitted response, depends on covariates (i.e., explanatory predictor variables) we can estimate regression models for hazard and for conditional accuracy.
Such covariates can be constant over within-trial time (e.g., gender, race, trial number, block number) or vary with within-trial time (e.g., heart rate, eye gaze position, eye pupil dilation). Note that time-varying covariates are not covered in this tutorial.

Heterogeneity

fitting problems

-------------------

@@

## 2.1 Basic features of event history analysis

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a discrimination task. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA of the RT data (hazard and survivor functions) can be extended with a discrete-time SAT analysis of the timed accuracy data. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

@@

In contrast to the large-*N* design that typically average across many participants without being able to scrutinize individual data patterns,  small-*N*  designs  retain  crucial  information  about  the  data  patterns of individual observers. This can be advantageous whenever participants differ systematically in their strategies or in the time courses of their effects, so that averaging them would lead to misleading data patterns. Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-*N* designs can still achieve what are generally considered acceptable levels of statistical power, if they have a sufficient amount of data overall [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018].

# 3. An overview of the general analytical workflow

@@

## 3.3 Implementation

>>>>

Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (< 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (>500 ms) are controlled by the target stimulus.

@@

(ref:descr-fig1-caption) Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus whose presence (s)he has to detect by pressing a button within a certain time window (e.g., the first second after target onset). In EHA parlance, the single event of interest is a button press response, *time zero* is defined as target display onset, the  *observation period* is 1 second long in each trial or repeated measurement, in each trial the participant is *at risk* for response occurrence as long as the response has not occurred yet, and the individual always starts in an "idle" state in each trial and *transitions* to a "detected" state when a response occurs.

@@

After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins; Figure 1B), let RT be a discrete random variable denoting the *rank* of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the detection response in trial 1 might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms). Thus, the RT data are interval-censored, because we only use the information that a<RT<=b when the response occurs in time bin (a,b].

While experimental psychologists are familiar with the cumulative distribution function or F(t) = P(RT <= t) and the probability mass function or P(t) = P(RT = t), discrete-time EHA focuses on the discrete-time hazard function of event occurrence:

\noindent h(t) = P(RT = t| RT $\geq$ t)   \hfill  (1)

\noindent and the discrete-time survivor function:

\noindent S(t) = P(RT $>$ t) = 1 - F(t) =  [1-h(t)].[1-h(t-1)].[1-h(t-2)]...[1-h(1)]  \hfill  (2)

The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. 
In other words, it reflects the instantaneous risk that the response occurs in bin t, given that it has not yet occurred in one of the prior bins. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the survival probability, the probability that the event does not occur before the endpoint of bin t. As a result, only the hazard function conveys the risk of event occurrence associated with each bin, and ... suited for online tracking of performance.. cfr mouse cursor movements....

For two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function

\noindent ca(t) = P(correct | RT = t)   \hfill  (5)

\noindent which gives you for each bin the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. The ca(t) function is also known as the micro-level speed-accuracy tradeoff (SAT) function.
We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

As we will illustrate in Tutorials 1a and 1b, performing a descriptive EHA/SAT analysis by calculating the sample-based estimates of h(t), S(t) and ca(t) for each combination of participant and condition requires setting up a *life table*.
Definition life table...

## 2.5 Bayesian vs. frequentist approaches to regression

To study how the risk of a response, and the accuracy of an emitted response, depends on covariates (i.e., explanatory predictor variables) we can estimate regression models for hazard and for conditional accuracy, i.e., perform inferential EHA/SAT analysis. Such covariates can be constant over within-trial time (e.g., gender, race, trial number, block number) or vary with within-trial time (e.g., heart rate, eye gaze position, eye pupil dilation). Note that time-varying covariates are not covered in this tutorial.

Heterogeneity -> Multilevel survival analysis: Methods, Models and Applications
Austin 2017 !!

fitting problems -> Bayesian

## 2.6 Number of samples, repeated measures, time bins

In a typical RT data set from a within-subject design, there are N individuals and M repeated measures or trials per experimental condition.
To test process models of cognition, .. advises  to use small-N designs, ... eACH SUBJECT REPLICATION UNIT

Power IS A COMPLEX FUNCTION OF .... (REF A, REF B ON POWER WITH EXPONENTIAL)

@@

bin width
-------------------

@@

<!--## 2.1 Basic features of event history analysis

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a discrimination task. For example, a task may involve choosing between two response options with only one of them being correct.

@@

In contrast to the large-*N* design that typically average across many participants without being able to scrutinize individual data patterns,  small-*N*  designs  retain  crucial  information  about  the  data  patterns of individual observers. This can be advantageous whenever participants differ systematically in their strategies or in the time courses of their effects, so that averaging them would lead to misleading data patterns. Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-*N* designs can still achieve what are generally considered acceptable levels of statistical power, if they have a sufficient amount of data overall [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018].
-->

<!--# 3. An overview of the general analytical workflow

@@

## 3.3 Implementation-->

### 2025-06-18 manuscript/manuscript.tex
add section 2.4

@@

\caption{Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100{]}, the last bin is (900,1000{]}. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B). Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (\textless{} 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (\textgreater500 ms) are controlled by the target stimulus.}\label{fig:plot1}
\end{figure}

@@

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus whose presence (s)he has to detect by pressing a button within a certain time window (e.g., the first second after target onset). In EHA parlance, the single event of interest is a button press response, \emph{time zero} is defined as target display onset, the \emph{observation period} is 1 second long in each trial or repeated measurement, in each trial the participant is \emph{at risk} for response occurrence as long as the response has not occurred yet, and the individual starts in an ``idle'' state in each trial and \emph{transitions} to a ``detected'' state when a response occurs.

@@

However, in this tutorial we focus on discrete-time methods for three reasons:
First, we are interested in studying the shape of the hazard function (Cox regression ignores this and only tests the effects of covariates);
Second, empirical hazard and conditional accuracy functions from certain RT tasks (e.g., interference tasks; Figure 1B) can show abrupt changes in their shape (parametric methods assume smooth distributions), and the shape of the hazard function in many experimental tasks is still unknown (parametric methods assume well-defined probability distributions);
Third, in discrete time, hazard is simply defined as a conditional probability (see 2.4) and we can apply logistic regression modeling with which most experimental psychologists are familiar.

Thus, we believe that discrete-time methods are a good starting point for experimental psychologists that want to abandon ANOVA, even though continuous-time methods might be preferred in some situations.

@@

After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins), let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the first response might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).

Interval-censored\ldots{} +

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence

and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in section A of the Supplemental Material.

def hazard

def survival

def cumulative distribution

def prob mass function

def conditional accuracz function

The discrete-time hazard function for individual i in trial j of condition k gives you, for each time bin, the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins.

In other words, it reflects the instantaneous risk that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the survival probability, the probability that the event occurs after bin t. In other words, the survivor function gives you for each time bin the likelihood that the event occurs in the future, i.e., in one of the subsequent time bins.

life table\ldots{}

\subsection{2.5 Number of samples, repeated measures, time bins}\label{number-of-samples-repeated-measures-time-bins}

In a typical RT data set from a within-subject design, there are N individuals and M repeated measures or trials per experimental condition.

@@

\subsection{2.6 Bayesian vs.~frequentist approach}\label{bayesian-vs.-frequentist-approach}

To study how the risk of a response, and the accuracy of an emitted response, depends on covariates (i.e., explanatory predictor variables) we can estimate regression models for hazard and for conditional accuracy.
Such covariates can be constant over within-trial time (e.g., gender, race, trial number, block number) or vary with within-trial time (e.g., heart rate, eye gaze position, eye pupil dilation). Note that time-varying covariates are not used in this tutorial.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

@@

\subsection{2.1 Basic features of event history analysis}\label{basic-features-of-event-history-analysis}

In EHA, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.

Since our focus here is on hazard models that use discrete time units, we describe that approach.

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../sims/figures/haz_surv_single}

\caption{Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. Error bars represent +/- 1 standard error of the respective proportion. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t=5) = .44.
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.}\label{fig:plot2}
\end{figure}

\subsection{2.3 Event history analysis in the context of experimental psychology}\label{event-history-analysis-in-the-context-of-experimental-psychology}

To make EHA more relevant to researchers studying cognitive psychology and cognitive neuroscience, in this section we provide a relevant worked example and consider implications that are relevant to that domain of research.

\subsubsection{2.3.1 A worked example}\label{a-worked-example}

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a discrimination task. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA of the RT data (hazard and survivor functions) can be extended with a discrete-time SAT analysis of the timed accuracy data. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977). We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical choice RT example that is inspired by real data (Panis \& Schmidt, 2016), but simplified to make the main point clearer (Figure 3). In a standard priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target.

Figure 3 shows that the early upswing in hazard is equal for both priming conditions (Figure 3, upper panel), and that early emitted responses are always correct in the congruent condition and always incorrect in the incongruent condition (Figure 3, lower panel). These results show that for short waiting times (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). During time bin 6 the target-triggered response channel is activated and causes response competition -- ca(6) = .5 -- and a lower hazard probability in the incongruent condition. For waiting times of 600 ms or more, the hazard of response occurrence is lower in incongruent compared to congruent trials, and all responses emitted in these late bins are correct.

This joint pattern of results is interesting because it can provide meaningfully different conclusions about psychological processes compared to conventional analyses, such as computing mean-average RT and accuracy across trials. Mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. For instance, if mean-average RT was higher in incongruent than congruent trials, one may conclude that cognitive mechanisms that support interference control are working as expected across trials, and are indexed by each recorded response. But such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

Unlocking the temporal states of cognitive processes can be revealing for theory development and the understanding of basic psychological processes. Possibly more importantly, however, is that it simultaneously opens the door to address many new and previously unanswered questions. Do all participants show similar temporal states or are there individual differences? Do such individual differences extend to those individuals that have been diagnosed with some form of psychopathology? How do temporal states relate to brain-based mechanisms that might be studied using other methods from cognitive neuroscience? And how much of theory in cognitive psychology would be in need of revision if mean-average comparisons were supplemented with a temporal states approach?

\subsubsection{2.3.2 Implications for designing experiments}\label{implications-for-designing-experiments}

Performing EHA in experimental psychology has implications for how experiments are designed. Indeed, if trials are categorised as a function of when responses occur, then each time bin will only include a subset of the total number of trials. For example, let's consider an experiment where each participant performs 2 conditions and there are 100 trial repetitions per condition. Those 100 trials must be distributed in some manner across the chosen number of bins.

In such experimental designs, since the number of trials per condition are spread across bins, it is important to have a relatively large number of trial repetitions per participant and per condition. Accordingly, experimental designs using this approach typically focus on factorial, within-subject designs, in which a large number of observations are made on a relatively small number of participants (so-called small-\emph{N} designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design (Baker et al., 2021; Smith \& Little, 2018).

In contrast to the large-\emph{N} design that typically average across many participants without being able to scrutinize individual data patterns, small-\emph{N} designs retain crucial information about the data patterns of individual observers. This can be advantageous whenever participants differ systematically in their strategies or in the time courses of their effects, so that averaging them would lead to misleading data patterns. Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-\emph{N} designs can still achieve what are generally considered acceptable levels of statistical power, if they have a sufficient amount of data overall (Baker et al., 2021; Smith \& Little, 2018).

\section{3. An overview of the general analytical workflow}\label{an-overview-of-the-general-analytical-workflow}

Although the focus is on EHA/SAT, we also want to briefly comment on broader aspects of our general analytical workflow, which relate more to data science and data analysis workflows.

\subsection{3.1 Data science workflow and descriptive statistics}\label{data-science-workflow-and-descriptive-statistics}

We perform data wrangling following tidyverse principles and a functional programming approach (Wickham, Çetinkaya-Rundel, \& Grolemund, 2023).
In short, functional programming means that you avoid writing your own loops and instead use functions that have been built and tested by others.
In addition, we also supply a set of custom-built functions, which make the process of data wrangling in the context of data preparation and descriptive statistics a lot quicker and more efficient.

\subsection{3.2 Inferential statistical approach}\label{inferential-statistical-approach}

Our lab adopts an estimation approach to multilevel regression (Kruschke \& Liddell, 2018; Winter, 2019), which is heavily influenced by the Bayesian framework as suggested by Richard McElreath (Kurz, 2023b; McElreath, 2020). We also use a ``keep it maximal'' approach to specifying varying (or random) effects (Barr, Levy, Scheepers, \& Tily, 2013). This means that wherever possible we include varying intercepts and slopes per participant.
To make inferences, we use two main approaches. We compare models of different complexity, using information criteria (e.g., WAIC) and cross-validation (e.g., LOO), to evaluate out-of-sample predictive accuracy (McElreath, 2020). We also take the most complex model and evaluate key parameters of interest using point and interval estimates.

\subsection{3.3 Implementation}\label{implementation}

@@

\bibitem[\citeproctext]{ref-bakerPowerContoursOptimising2021}
Baker, D. H., Vilidaite, G., Lygo, F. A., Smith, A. K., Flack, T. R., Gouws, A. D., \& Andrews, T. J. (2021). Power contours: {Optimising} sample size and precision in experimental psychology and human neuroscience. \emph{Psychological Methods}, \emph{26}(3), 295--314. \url{https://doi.org/10.1037/met0000337}

\bibitem[\citeproctext]{ref-barrRandomEffectsStructure2013}
Barr, D. J., Levy, R., Scheepers, C., \& Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: {Keep} it maximal. \emph{Journal of Memory and Language}, \emph{68}(3), 10.1016/j.jml.2012.11.001. \url{https://doi.org/10.1016/j.jml.2012.11.001}

@@

\bibitem[\citeproctext]{ref-kruschkeBayesianNewStatistics2018}
Kruschke, J. K., \& Liddell, T. M. (2018). The {Bayesian New Statistics}: {Hypothesis} testing, estimation, meta-analysis, and power analysis from a {Bayesian} perspective. \emph{Psychonomic Bulletin \& Review}, \emph{25}(1), 178--206. \url{https://doi.org/10.3758/s13423-016-1221-4}

@@

\bibitem[\citeproctext]{ref-smithSmallBeautifulDefense2018}
Smith, P. L., \& Little, D. R. (2018). Small is beautiful: {In} defense of the small-{N} design. \emph{Psychonomic Bulletin \& Review}, \emph{25}(6), 2083--2101. \url{https://doi.org/10.3758/s13423-018-1451-8}

@@

\bibitem[\citeproctext]{ref-wickhamDataScienceImport2023}
Wickham, H., Çetinkaya-Rundel, M., \& Grolemund, G. (2023). \emph{R for data science: Import, tidy, transform, visualize, and model data} (2nd edition). Beijing Boston Farnham Sebastopol Tokyo: O'Reilly.

>>>>

Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (\textless{} 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (\textgreater500 ms) are controlled by the target stimulus.

@@

\caption{Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100{]}, the last bin is (900,1000{]}. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).}\label{fig:plot1}
\end{figure}

@@

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus whose presence (s)he has to detect by pressing a button within a certain time window (e.g., the first second after target onset). In EHA parlance, the single event of interest is a button press response, \emph{time zero} is defined as target display onset, the \emph{observation period} is 1 second long in each trial or repeated measurement, in each trial the participant is \emph{at risk} for response occurrence as long as the response has not occurred yet, and the individual always starts in an ``idle'' state in each trial and \emph{transitions} to a ``detected'' state when a response occurs.

@@

However, in this tutorial we focus on discrete-time methods for three reasons:
First, we are interested in studying the shape of the hazard function (Cox regression ignores this and only tests the effects of covariates);
Second, empirical hazard and conditional accuracy functions from certain RT tasks (e.g., interference tasks; Figure 1B) can show abrupt changes in their shape (parametric methods assume smooth distributions), and the shape of the hazard function in many experimental tasks is still unknown (parametric methods assume well-defined probability distributions);
Third, in discrete time, hazard is simply defined as a conditional probability (see 2.4) and we can apply logistic regression modeling with which most experimental psychologists are already familiar.

In sum, due to their simplicity and flexibility, we believe that discrete-time methods are a good starting point for experimental psychologists that want to abandon ANOVA and learn to apply EHA, even though continuous-time methods might be more suited in certain situations.

@@

After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins; Figure 1B), let RT be a discrete random variable denoting the \emph{rank} of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the detection response in trial 1 might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms). Thus, the RT data are interval-censored, because we only use the information that a\textless RT\textless=b when the response occurs in time bin (a,b{]}.

While experimental psychologists are familiar with the cumulative distribution function or F(t) = P(RT \textless= t) and the probability mass function or P(t) = P(RT = t), discrete-time EHA focuses on the discrete-time hazard function of event occurrence:

\noindent h(t) = P(RT = t\textbar{} RT \(\geq\) t) \hfill  (1)

\noindent and the discrete-time survivor function:

\noindent S(t) = P(RT \(>\) t) = 1 - F(t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]} \hfill  (2)

The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins.
In other words, it reflects the instantaneous risk that the response occurs in bin t, given that it has not yet occurred in one of the prior bins. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the survival probability, the probability that the event does not occur before the endpoint of bin t. As a result, only the hazard function conveys the risk of event occurrence associated with each bin, and \ldots{} suited for online tracking of performance.. cfr mouse cursor movements\ldots.

For two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function

\noindent ca(t) = P(correct \textbar{} RT = t) \hfill  (5)

\noindent which gives you for each bin the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977). The ca(t) function is also known as the micro-level speed-accuracy tradeoff (SAT) function.
We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

As we will illustrate in Tutorials 1a and 1b, performing a descriptive EHA/SAT analysis by calculating the sample-based estimates of h(t), S(t) and ca(t) for each combination of participant and condition requires setting up a \emph{life table}.
Definition life table\ldots{}

\subsection{2.5 Bayesian vs.~frequentist approaches to regression}\label{bayesian-vs.-frequentist-approaches-to-regression}

To study how the risk of a response, and the accuracy of an emitted response, depends on covariates (i.e., explanatory predictor variables) we can estimate regression models for hazard and for conditional accuracy. Such covariates can be constant over within-trial time (e.g., gender, race, trial number, block number) or vary with within-trial time (e.g., heart rate, eye gaze position, eye pupil dilation). Note that time-varying covariates are not covered in this tutorial.

Multilevel survival analysis: Methods, Models and Applications
Austin 2017 !!

\subsection{2.6 Number of samples, repeated measures, time bins}\label{number-of-samples-repeated-measures-time-bins}

In a typical RT data set from a within-subject design, there are N individuals and M repeated measures or trials per experimental condition.
To test process models of cognition, researchers tend to use small-N designs, \ldots{}

@@

\subsection{bin width}\label{bin-width}

@@

@@

@@

@@

@@
